---
title: "Unsupervised Flow Data Analysis kinetics"
author: "TheVeger"
date: "14/07/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = ".")
```

# Libraries
```{r libs, message = FALSE}
library(tidyverse)
library(SingleCellExperiment)
library(CATALYST)
library(readxl)
library(clustree)
library(Rphenograph)
library(MetBrewer)
library(ggpubr)
library(svglite)
library(pals)
library(FlowSOM)
library(ConsensusClusterPlus)
library(flowCore)
library(ggrepel)
library(scater)
library(rasterpdf)
library(cowplot)
library(ComplexHeatmap)
library(diffcyt) #quantification of single cell data
library(rstatix)
library(hexbin)
library(ggrastr)
```

# Set paths
```{r set paths, warning=FALSE}
rm(list = ls())

path.data <- "../Data/"
path.files <- "../R_files/"
path.out <- "../R_out/"

H <- met.brewer("Kandinsky",n=20)[c(16)]
d1 <- met.brewer("Archambault",n=7)[c(3)]
d3 <- met.brewer("Kandinsky",n=20)[c(2)]
d7 <- met.brewer("Kandinsky",n=20)[c(7)]
d14 <- met.brewer("Archambault",n=7)[c(4)]
d21 <- met.brewer("Archambault",n=7)[c(5)]
d28 <- met.brewer("Archambault",n=20)[c(2)]
colvec <- c(H, d1, d3, d7, d14, d21, d28)
source("flow_functions.R")
```

# Prepare SCE
## Import SCE object
Import SCE object that we have saved previously.
```{r load and coldata}
sce <- readRDS(file = paste0(path.files, "01_SCE.rds"))
```

# Add marker_classes to rowdata
```{r rowdata}
rownames(sce) 
lineage_markers <- c("CD11c", "TCRb chain", "CD4", "gdTCR","F4_80", "CD90.2", "CD11b", "CD127", "MHC-II", "Ly-6C", "Ly-6G", "CD19", "NK1.1", "CD8") #lineage markers used for clustering

rowData(sce)[rowData(sce)$marker_name %in% lineage_markers, "marker_class"] <- "type"
rowData(sce)[!rowData(sce)$marker_name %in% lineage_markers, "marker_class"] <- "state"
rowData(sce)[rowData(sce)$marker_name %in% c("Live_Dead","CD45"), "marker_class"] <- "none" 

type_markers(sce)
state_markers(sce)
```

# Exploratory Data Analysis
```{r Health check}
set.seed(1234)
# use cells= ">total amount of cells" if you want to use all cells
sce <- run_DR(sce, dr = "UMAP", markers = type_markers(sce), cells=10000000, scale = TRUE)
plotDR(sce, "UMAP", color_by = "timepoint")+
   scale_fill_manual(values = colvec)+
  scale_color_manual(values = colvec)
plotDR(sce, "UMAP", color_by = c(type_markers(sce)), ncol = round(sqrt(16*length(c(type_markers(sce)))/9)))
plotDR(sce, "UMAP", color_by = c(state_markers(sce)), ncol = round(sqrt(16*length(c(state_markers(sce)))/9)))
```
# Primary 
## flowsom clustering
Here, we perform the primary clustering and dimensionality reduction (UMAP). The "cluster" function of CATALYST runs a FlowSOM clustering followed by a Consensus Meta clustering which is a standard clustering method and performs clusterings very rapidly.
```{r primary clustering}
kinit <- 30 #try 10 or 20

sce <- flowsom(sce, markers = type_markers(sce), maxK = kinit, scale = TRUE, seed = 1234) 
plot_DR(sce, color_by = "flowsom")
plot_DR(sce, color_by = type_markers(sce))

#Check the validity of the cluster: If your density plots show bimodal distributions, you may be underclustering.
plotClusterExprs(sce, k = paste0("meta", kinit), features = "type") + labs(title = "primary clustering marker expression histogram")

#If all density plots show one peak, it is valid to look at the heatmap. Else, this will hide information.
plotExprHeatmap(sce, features = "type",
    by = "cluster_id", k = paste0("meta", kinit),
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)
saveRDS(object = sce, file = paste0(path.files, "01_SCE_flowsom.rds"))
#save SCE in between steps to continue at different steps of the script.
sce <- readRDS(file = paste0(path.files, "01_SCE_flowsom.rds"))
```

## Merge primary clustering
Label and merge clusters
```{r merge primary clusters}
(merging_table <- read_excel(paste0(path.files, "merged_clusters_primary.xlsx")))
merging_table <- merging_table[,1:2]


sce <- mergeClusters(sce, k = paste0("meta", kinit), table = merging_table, id = "primary", overwrite = TRUE)

# Add merged cluster ids to colData
code <- cluster_codes(sce)[,c("som100", "primary")]
coldata <- as.data.frame(colData(sce))
coldata <- merge(coldata, code, by.x = "cluster_id", by.y = "som100")
colData(sce) <- DataFrame(coldata[order(coldata$cell_id),])


p1 <- plot_DR(sce, "UMAP", color_by = "primary")+
  theme(legend.text = element_text(size=14),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        legend.title = element_text(size=18, face = "bold"))

pdf(paste0(path.out, "2a_merged_clustering.pdf"), height = 9, width = 11, onefile = TRUE)
p1
features <- plot_DR(sce, "UMAP", color_by = type_markers(sce), ncol = 5)+
  theme(strip.text = element_text(size=18),
        panel.grid = element_blank(),
        strip.background = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank(),
        legend.text =   element_text(size=14),
        legend.title = element_text(size=18, face = "bold"))
plotExprHeatmap(sce, features = "type",
    by = "cluster_id", k = "primary",
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)
dev.off()

```
# Save primary SCE object
```{r save RDS}
saveRDS(object = sce, file = paste0(path.files, "SCE_primary_clustered.rds"))
```

# Secondary clustering on T-like lymphocytes
```{r}
sce <- readRDS(file = paste0(path.files, "SCE_primary_clustered.rds"))
sce <- filterSCE(sce, primary %in% c("abT Cells",  "gdT Cells", "ILCs")) #filter for secondary clustering
sce$cluster_id
#lineage markers used for secondary clustering
lineage_markers <- c("TCRb chain", "CD4", "gdTCR", "CD90.2", "CD127","RORgt", "CD8", "GATA3", "FOXP3", "CD25", "CD122", "T-bet") 

rowData(sce)[rowData(sce)$marker_name %in% lineage_markers, "marker_class"] <- "type"
rowData(sce)[!rowData(sce)$marker_name %in% lineage_markers, "marker_class"] <- "none"
type_markers(sce)
sce <- run_DR(sce, dr = "UMAP", markers = type_markers(sce), cells = 10000000, scale = TRUE)
sce <- phenograph(sce, markers = type_markers(sce), assay = "exprs", k = 11, scale = TRUE) #lower k is more clusters

ggsave(file=paste0(path.out,"Secondary_clustering.pdf"), plot=
         plot_DR(sce, color_by = "pheno", add_labels = TRUE),
       width=11, height=9)
ggsave(file=paste0(path.out,"Secondary_clustering_features.pdf"), plot=
         plot_DR(sce, color_by = type_markers(sce)), 
       width=11, height=9)


cluster_codes(sce) #code for the clusters
pdf(paste0(path.out, "Secondary_clustering_heatmap.pdf"), height = 9, width = 11, onefile = TRUE)
plotExprHeatmap(sce, features = "type", #plot heatmap for Pheno clustering
    by = "cluster_id", k = "custom",
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)

plotClusterExprs(sce, k = "custom", features = "type") + labs(title = "secondary clustering marker expression histogram")
dev.off()
```

#Merging secondary clustering
```{r}
(merging_table <- read_excel(paste0(path.files, "merged_clusters_secondaryT.xlsx")))
merging_table <- merging_table[,1:2]                   

sce <- mergeClusters(sce, k = "custom", table = merging_table, id = "secondary", overwrite = TRUE)

# Add merged cluster ids to colData
code <- cluster_codes(sce)[,c("custom", "secondary")]
coldata <- as.data.frame(colData(sce))
coldata <- merge(coldata, code, by.x = "pheno", by.y = "custom")
colData(sce) <- DataFrame(coldata[order(coldata$cell_id),])

saveRDS(object = sce, file = paste0(path.files, "SCE_secondaryT_clustered.rds"))
sce <- readRDS(file = paste0(path.files, "SCE_secondaryT_clustered.rds"))

p1 <- plot_DR(sce, "UMAP", color_by = "secondary")+
  theme(legend.text = element_text(size=14),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        legend.title = element_text(size=18, face = "bold"))

pdf(paste0(path.out, "merged_secondaryT.pdf"), height = 9, width = 11, onefile = TRUE)
p1
plot_DR(sce, "UMAP", color_by = type_markers(sce), ncol = 4)+
         theme(strip.text = element_text(size=18),
               panel.grid = element_blank(),
               strip.background = element_blank(),
               axis.title = element_blank(),
               axis.text = element_blank(),
               legend.text =   element_text(size=14),
               legend.title = element_text(size=18, face = "bold"))
plotExprHeatmap(sce, features = "type",
    by = "cluster_id", k = "secondary",
    scale = "first", q = 0.01, perc = TRUE, col_dend = FALSE, bars = TRUE, col_clust = FALSE)
dev.off()
```


# Merging the primary and secondary clustering
```{r}
sce_prim <- readRDS(file = paste0(path.files, "SCE_primary_clustered.rds"))
sce_secT <- readRDS(file = paste0(path.files, "SCE_secondaryT_clustered.rds"))

prim_cluster_id <- as.data.frame(colData(sce_prim)[,c("cell_id", "primary")]) %>% 
  dplyr::rename(cluster = primary)
Lymcell_cluster_id <- as.data.frame(colData(sce_secT)[,c("cell_id", "secondary")]) %>% 
  dplyr::rename(cluster = secondary)
subcluster_id <- prim_cluster_id %>%
  full_join(Lymcell_cluster_id, by = "cell_id") %>% 
  mutate(cluster.x = as.character(cluster.x)) %>% 
  mutate(cluster.y = as.character(cluster.y)) %>% 
  mutate(cluster = ifelse(is.na(cluster.y), cluster.x, cluster.y)) %>%
  # Drop the unnecessary columns
  select(-cluster.y) %>% 
  dplyr::rename(original = cluster.x)

coldata <- data.frame(colData(sce_prim)) %>% dplyr::select(-c("flowsom", "primary"))
me <- merge(coldata, subcluster_id, by = c("cell_id"), all = TRUE)
me$cluster <- factor(me$cluster)
colData(sce_prim) <- DataFrame(me)
metadata(sce_prim)$cluster_codes <- data.frame(
    custom = factor(levels(me$cluster_id), levels = levels(me$cluster_id)))

saveRDS(object = sce_prim, file = paste0(path.files, "SCE_all_merged.rds"))
sce <- readRDS(file = paste0(path.files, "SCE_all_merged.rds"))
p1 <- plot_DR(sce, "UMAP", color_by = "cluster")+
  theme(legend.text = element_text(size=14),
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        legend.title = element_text(size=18, face = "bold"))+
  guides(color = guide_legend(ncol = 1, override.aes = list(size = 3)))

pdf(paste0(path.out, "all_clustering_combined.pdf"), height = 6, width = 9, onefile = TRUE)
p1
dev.off()

```

```{r save RDS}
saveRDS(object = sce_prim, file = paste0(path.files, "SCE_all_merged.rds"))
```

# Session Info
In the session info you find all the packages and their version that were used. This is helpful for other people that want to reproduce your code and for you if you want to cite a package.
```{r}
sessionInfo()
```

