---
title: "Plots of Unsupervised Flow Data Analysis Kinetics"
author: "TheVeger"
date: "14/07/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = ".")
```

# Libraries
```{r libs, message = FALSE}
library(tidyverse)
library(SingleCellExperiment)
library(readxl)
library(clustree)
library(MetBrewer)
library(ggpubr)
library(rstatix) #do statistics for plots
library(pals)
library(openxlsx)
library(CATALYST)
library(ConsensusClusterPlus)
library(flowCore)
library(ggrepel)
library(scater)
library(rasterpdf)
library(scales) #add percentages
```

# Set paths
```{r set paths, warning=FALSE}
rm(list = ls())

path.data <- "../Data/"
path.files <- "../R_files/"
path.out <- "../R_out/"

H <- met.brewer("Kandinsky",n=20)[c(16)]
d1 <- met.brewer("Archambault",n=7)[c(3)]
d3 <- met.brewer("Kandinsky",n=20)[c(2)]
d7 <- met.brewer("Kandinsky",n=20)[c(7)]
d14 <- met.brewer("Archambault",n=7)[c(4)]
d21 <- met.brewer("Archambault",n=7)[c(5)]
d28 <- met.brewer("Archambault",n=20)[c(2)]

colvec <- c(H, d1, d3, d7, d14, d21, d28)

Treatment <- met.brewer("Navajo",n=20)[c(15)]
Treatment2 <- met.brewer("Navajo",n=20)[c(1)]
source("flow_functions.R")
```

# Prepare SCE
## Import SCE object

Import SCE object that we have saved previously.
```{r load and coldata}
sce <- readRDS(file = paste0(path.files, "SCE_all_merged.rds"))
```

## DA visualisation

# Data preperation
```{r DA}
sce$timepoint <- factor(sce$timepoint, levels = c("H", "d1", "d3","d7", "d14", "d21", "d29"))

```

# Density
```{r}
pdf(paste0(path.out, "DensityPlots.pdf"), height = 6, width = 6, onefile = TRUE)
plot_density(x = sce, color_by = "timepoint", ncol =2, colvec = colvec, bins = 12, xlim = c(-13, 13.5), ylim = c(-17, 18), contures = 1)+
  theme(legend.position = "none",
        axis.text.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title = element_blank(),
        axis.ticks = element_blank(),
        strip.text = element_blank())
dev.off()
```

# LiveCD45 plots
```{r}
LiveCD45 <- readRDS(paste0(path.files, "LiveCD45Counts.rds"))
LiveCD45 <- LiveCD45 %>% dplyr::filter(!((sample_id == "Bladder_CTRL_d1_2024_01_24_5" | sample_id == "Bladder_CTRL_d1_2024_01_24_7")))
LiveCD45$Timepoint <- factor(LiveCD45$Timepoint, levels = c("H", "d1", "d3","d7", "d14", "d21", "d29"))
LiveCD45 <- LiveCD45 %>% mutate(LogMeasurement = log(LiveCD45))
LiveCD45 <- LiveCD45 %>% mutate("% of Single Cells" = LiveCD45/`Single Cells`)

stat.test <- LiveCD45 %>%
    group_by("Timepoint") %>%
    t_test(LiveCD45 ~ Timepoint, ref.group = "H") %>%
    add_significance("p")
  stat.test
  stat.test <- stat.test %>%
    add_xy_position(x = "Timepoint", dodge = 0.75, fun = "max", scales = "free", step.increase = 0)%>%
    dplyr::filter(p != "NaN") %>% mutate("y.position" = y.position*1.05)

pdf(paste0(path.out,format(Sys.Date(), "%Y%m%d"),"_","LiveCD45Count.pdf"), onefile = TRUE, width = 5.333, height = 3)
ggplot(LiveCD45, aes(x = Timepoint, y = LiveCD45, interaction(Timepoint)))+
    geom_boxplot(color = "black", fill = Treatment2, alpha = 0.5, width = 0.25, outlier.shape = NA)+
    stat_summary(fun=median, geom="line", aes(group=1), color = "black")  + 
    geom_point(color = Treatment2, fill = Treatment2,
               position = position_jitter(width = 0.1), # Jitter points
               alpha = 0.5, size = 2, shape = 16) +
    scale_y_log10(expand = expansion(mult = 0.1))+
    stat_pvalue_manual(stat.test,  label = "p", tip.length = 0.02, hide.ns = FALSE, remove.bracket = TRUE)+
    theme_bw()+
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_line(color = "black"),
          panel.spacing = unit(0, "lines"),
          strip.background = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size =18 , face= "bold", color = "black"),
          axis.text.y = element_text(size =13, color = "black"),
          axis.title.y = element_text(size =20 , face= "bold", color = "black"),
          legend.position = "none",
          strip.text = element_text(size = 18, face = "bold"))+
    labs(y = "Count LiveCD45")+
    scale_fill_manual(values = colvec)+
    scale_color_manual(values = colvec)
dev.off()

stat.test <- LiveCD45 %>%
    group_by("Timepoint") %>%
    t_test(`% of Single Cells` ~ Timepoint, ref.group = "H") %>%
    add_significance("p")
  stat.test
  stat.test <- stat.test %>%
    add_xy_position(x = "Timepoint", dodge = 0.75, fun = "max", scales = "free", step.increase = 0)%>%
    dplyr::filter(p != "NaN") %>% mutate("y.position" = y.position*1.05)

pdf(paste0(path.out,format(Sys.Date(), "%Y%m%d"),"_","LiveCD45PercentofSingleCells.pdf"), onefile = TRUE, width = 5.333, height = 3)
ggplot(LiveCD45, aes(x = Timepoint, y = `% of Single Cells`, interaction(Timepoint)))+
    geom_boxplot(color = "black", fill = Treatment, alpha = 0.5, width = 0.25, outlier.shape = NA)+
    stat_summary(fun=median, geom="line", aes(group=1), color = "black")  + 
    geom_point(color = Treatment, fill = Treatment,
               position = position_jitter(width = 0.1), # Jitter points
               alpha = 0.5, size = 2, shape = 16) +
    scale_y_continuous(labels = percent, expand = expansion(mult = 0.1)) +
    stat_pvalue_manual(stat.test,  label = "p", tip.length = 0.02, hide.ns = FALSE, remove.bracket = TRUE)+
    theme_bw()+
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_line(color = "black"),
          panel.spacing = unit(0, "lines"),
          strip.background = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size =18 , face= "bold", color = "black"),
          axis.text.y = element_text(size =18, color = "black"),
          axis.title.y = element_text(size =20 , face= "bold", color = "black"),
          legend.position = "none",
          strip.text = element_text(size = 18, face = "bold"))+
    labs(y = "LiveCD45 % of Single Cells")+
    scale_fill_manual(values = colvec)+
    scale_color_manual(values = colvec)
dev.off()
```
# Plotting single populations
```{r}
Fig1 <- calculate_DA(sce, cluster_colum = "cluster", sample_id = "sample_id", condition = NULL)
Fig1$Freq <- as.double(Fig1$Freq)
cellcounts <- readRDS(paste0(path.files,"cellcounts.rds"))
Fig1 <- Fig1 %>% left_join(cellcounts, by = "sample_id")
Fig1 <- Fig1 %>% mutate("count" = ((Freq*LiveCD45)/Count)*10^6*20*0.4)
# Neutrophils of d1 from 20240124 were removed, as they were all dead. This is likely due to the new batch of Liberase Tm that I used in the first timepoints, which was much stronger that the previous lot. Therefore, at d21 p.i. of the second repitition of the kinetics, I changed the concentration of Liberase Tm. 
Fig1 <- Fig1 %>% dplyr::filter(!((sample_id == "Bladder_CTRL_d1_2024_01_24_5" | sample_id == "Bladder_CTRL_d1_2024_01_24_7") & cluster == "Neutrophils"))
write.xlsx(x = Fig1, file = paste0(path.files, "Fig1.xlsx"), sheetName = "data", rownames = FALSE)
Th <- Fig1[grepl("Th|Treg", Fig1$cluster), ]
Th <- Th %>%
  group_by(sample_id, cluster) %>%
  mutate(total_counts = sum(counts)) %>%
  ungroup() %>%
  group_by(sample_id) %>%
  mutate(cluster_sum = sum(counts)) %>%
  ungroup() %>%
  mutate(percentage = (counts / cluster_sum)) %>%
  select(-total_counts, -cluster_sum)

ILC <- Fig1[grepl("ILC", Fig1$cluster), ]
ILC <- ILC %>%
  group_by(sample_id, cluster) %>%
  mutate(total_counts = sum(counts)) %>%
  ungroup() %>%
  group_by(sample_id) %>%
  mutate(cluster_sum = sum(counts)) %>%
  ungroup() %>%
  mutate(percentage = (counts / cluster_sum)) %>%
  select(-total_counts, -cluster_sum)
gdT <- Fig1[grepl("gdT", Fig1$cluster), ]
gdT <- gdT %>%
  group_by(sample_id, cluster) %>%
  mutate(total_counts = sum(counts)) %>%
  ungroup() %>%
  group_by(sample_id) %>%
  mutate(cluster_sum = sum(counts)) %>%
  ungroup() %>%
  mutate(percentage = (counts / cluster_sum)) %>%
  select(-total_counts, -cluster_sum)

cd4_data <- Fig1 %>%
  filter(cluster %in% c("Th1", "Th2", "Th17", "Treg","Unpolarized CD4 T Cells"))

# Group by sample_id and summarize to create new total count and Freq for CD4 T Cells
cd4_summarized <- cd4_data %>%
  group_by(condition, timepoint, number, sample_id) %>%
  summarize(
    count = sum(count),
    Freq = sum(Freq)
  ) %>% mutate(cluster = "CD4 T Cells")
```

# Percentages
```{r}
pdf(paste0(path.out,format(Sys.Date(), "%Y%m%d"),"_","clusterpercentages.pdf"), onefile = TRUE, width = 5.333, height = 3)
for(t in unique(Fig1$cluster)){
  int <- Fig1 %>% dplyr::filter(cluster == t) #filter for measurements of the current type
  
  k <- int$cluster %>% unique() %>% length()
  
  # With the stat.test table we calculate the significance values, which we can later add to the plot.
  # Watch out that the T-test can only be performed if there's n>1 measurement per condition.
  stat.test <- int %>%
    group_by("cluster", "timepoint") %>%
    t_test(Freq ~ timepoint, ref.group = "H") %>%
    add_significance("p")
  stat.test
  stat.test <- stat.test %>%
    add_xy_position(x = "timepoint", dodge = 0.75, fun = "max", scales = "free", step.increase = 0)%>%
    dplyr::filter(p != "NaN") %>% mutate("y.position" = y.position*1.05)
  
  p1 <-ggplot(int, aes(x = timepoint, y = Freq, interaction(timepoint)))+
    geom_boxplot(color = "black", fill = Treatment, alpha = 0.5, width = 0.25, outlier.shape = NA)+
    stat_summary(fun=median, geom="line", aes(group=1), color = "black")  + 
    geom_point(color = Treatment, fill = Treatment,
               position = position_jitter(width = 0.1), # Jitter points
               alpha = 0.5, size = 2, shape = 16) +
    scale_y_continuous(labels = percent, expand = expansion(mult = 0.1)) + # we expand the y-axis by 30% to leave space for significance , expand = expansion(mult = c(.05, .05))
    stat_pvalue_manual(stat.test,  label = "p.signif", tip.length = 0.02, hide.ns = "p", remove.bracket = TRUE, size = 8)+
    theme_bw()+
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_line(color = "black"),
          panel.spacing = unit(0, "lines"),
          strip.background = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size =18 , face= "bold", color = "black"),
          axis.text.y = element_text(size =18, color = "black"), #mac is 23
          axis.title.y = element_text(size =20 , face= "bold", color = "black"),
          legend.position = "none",
          strip.text = element_text(size = 18, face = "bold"))+
    labs(y = paste0("% of ", "LiveCD45"),title = t)
  print(p1) #print plot (has to be specified in a for-loop)
}
dev.off()
```
# Counts
```{r}
pdf(paste0(path.out,format(Sys.Date(), "%Y%m%d"),"_","clustercounts.pdf"), onefile = TRUE, width = 5.333, height = 3)
for(t in unique(Fig1$cluster)){
  int <- Fig1 %>% dplyr::filter(cluster == t) #filter for measurements of the current type
  
  k <- int$cluster %>% unique() %>% length()
  
  # With the stat.test table we calculate the significance values, which we can later add to the plot.
  # Watch out that the T-test can only be performed if there's n>1 measurement per condition.
  stat.test <- int %>%
    group_by("cluster", "timepoint") %>%
    t_test(count ~ timepoint, ref.group = "H") %>%
    #adjust_pvalue(method = "bonferroni") %>%
    add_significance("p")
  stat.test
  stat.test <- stat.test %>%
    add_xy_position(x = "timepoint", dodge = 0.75, fun = "max", scales = "free", step.increase = 0)%>%
    dplyr::filter(p != "NaN") %>% mutate("y.position" = y.position*1.05)
  
  p1 <-ggplot(int, aes(x = timepoint, y = count, interaction(timepoint)))+
    geom_boxplot(color = "black", fill = Treatment2, alpha = 0.5, width = 0.25, outlier.shape = NA)+
    stat_summary(fun=median, geom="line", aes(group=1), color = "black")  + 
    geom_point(color = Treatment2, fill = Treatment2,
               position = position_jitter(width = 0.1), # Jitter points
               alpha = 0.5, size = 2, shape = 16) +
    scale_y_log10(expand = expansion(mult = 0.1))+
    stat_pvalue_manual(stat.test,  label = "p.signif", tip.length = 0.02, hide.ns = "p", remove.bracket = TRUE, size = 8)+
    theme_bw()+
    theme(panel.grid = element_blank(),
          panel.border = element_blank(),
          axis.line = element_line(color = "black"),
          panel.spacing = unit(0, "lines"),
          strip.background = element_blank(),
          axis.title.x = element_blank(),
          axis.text.x = element_text(size =18 , face= "bold", color = "black"),
          axis.text.y = element_text(size =13, color = "black"),
          axis.title.y = element_text(size =20 , face= "bold", color = "black"),
          legend.position = "none",
          strip.text = element_text(size = 18, face = "bold"))+
   labs(y = "Count", title = t)
  print(p1) #print plot (has to be specified in a for-loop)
}
dev.off()
```

# Save SCE object
Finally, we save the clustered & merged SCE object in case we want to continue working with it later on.
```{r save RDS}
saveRDS(object = sce, file = paste0(path.files, "SCE_clustered_plotted.rds"))
```

# Session Info
In the session info you find all the packages and their version that were used. This is helpful for other people that want to reproduce your code and for you if you want to cite a package.
```{r}
sessionInfo()
```

